The Sloan Digital Sky Survey Spectroscopic Reduction Pipeline

### 2006-11-09 doc/paper/paper2d.tex
Update.
Still working, with references

%------------------------------------------------------------------------------
% SUBJECT HEADINGS
%------------------------------------------------------------------------------
\emph{Subject headings: }
???
\end{abstract}

@@

In this paper, we describe the details of the SDSS 2-d spectroscopic pipeline
(idlspec2d) including steps starting with processing of the raw CCD frames to final
flux calibration and spectral co-addition.  Downstream pipelines which
calculate redshifts, spectral types, emission and absorption line identification,
and velocity dispersions are described elsewhere.
As of writing the 2-d pipeline has been run successfully on all spectroscopic
data obtained in the course of the survey from February 28, 2000 (MJD 2451602) 
to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
releases EDR through DR5 (Stoughton et al ??, Abazajian et al ??, 
Adelman-McCarthy et al)
, SDSS obtained survey quality spectroscopic data on
1656 fiber pluggings (640 fibers each in 1575 unique plates).   
A total of 972,111 sky positions were targetted, with the following 
breakdown in the best matched spectral classification:  651920 galaxies, 
208054 stars, 112137 QSOs and 51300 sky spectra.

Mean exposure time per plate is 3914 seconds, and the median is 3302 seconds.
In Figure 1 we show three of the data quality diagnostics. This figure shows
the effective signal-to-noise ratio of each spectroscopic plug plate 
normalized to 3600 seconds exposure time, the median seeing in arcseconds,
and the RMS guide probe offset
in arcseconds, versus time in days after Feb 26, 2000 (MJD 2451600).

@@

In Figure 1, we show a top-view schematic of one the spectrographs, and
will detail the light path below starting at the slithead.
The output light from the fiber is collimated with a spherical mirror
(of radius of curvature 1264 mm), which is aligned by 3 motor mounts to
control focus, tip and tilt.  A left/right Hartmann mask 
covers the collimator to act as both a spectrograph shutter and traditional 
Shack-Hartmann mask for focus tests.  The collimator in spectrograph 2
was coated with an enhanced silver coating and delivers slightly higher
reflectivity compared to the aluminum coated collimator in spectrograph 1.
The collimated beam encounters a 40mm thick dichroic (beamsplitter) which
reflect light blueward of 6000\AA.  The blue reflected light is dispersed
with a 640 l/mm grism with zero deviation at 4960\AA.  The red transmitted
light is dispersed with a 440 l/mm grism with zero deviation at 7400\AA.
The dispersed light in each arm of the spectrograph enters a transmission camera
with a 240mm focal length (demagnification of 2.5) which is tuned for the 
corresponding range of wavelengths.  Each of the 4 cameras produce RMS 
spot sizes of approximately 20$\mu m$ over a 16.5$\degree$ diameter 
field-of-view delivered to a single thinned Tektronix 2048x2048x24$\mu m$ CCD
(see Gunn et al. 2006).
An external focus ring is adjustable by hand for each CCD camera to accomodate
differential focus changes not corrected with the common collimator.
The four spectrograph cameras are labeled $``$b1, r1, b2, r2", where
1 and 2 refer to the spectrographs and b and r to their blue and red sides.

@@

A plate is mounted in the telescope through a efficient procedure described
by Gunn et al (2006).  The switch from the imaging camera to a spectroscopic
cartridge takes an average of 30 minutes, while a switch from one cartridge
to another takes 10 minutes.  Once the cartridge is mounted, 
the telescope is slewed to the coordinates of the plate center.  
Before acquisition of the guide stars, the 8 flat-field petals are closed and
a sequence of one 10-second flat-field exposure and one 2-second 
Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
being read out, the flat-field petals are opened, and guide stars are acquired
through the 11 coherent guide bundles (9 are 7$"$ in diameter and 
2 are even larger at 11$"$).
With at least 4 guide stars acquired, closed loop guiding is started, and
usually converges to stable guiding within 2 minutes.
The plate scale of the SDSS telescope can be adjusted, and the guiding software
suggests a scale change if one is required.  Once close loop 
guiding has settled, the first science exposure is started.  The typical 
exposure time is 900 seconds, but exposure times from 600 to 1200 second are
not unexpected based on conditions and time constraints.  A second exposure
is started immediately after the first science exposure is complete and read
out.
A fast Linux workstation onsite copies the data from the data acquisition host,
and automatically performs a quick spectral extraction with a subset of the 
steps explained here.  The results are posted to sos.apo.nmsu.edu, and
the SDSS observers can check the signal-to-noise ratio (S/N) 
obtained in each of the 4 spectroscopic
cameras in real-time.  A plate is considered to have met SDSS survey standards
if all 4 cameras have a cumulative (S/N)$^2$ greater than 15 at 
fidicual magnitudes (what the heck is a fiducial magnitude??)
of 20.33 and 20.06 in SDSS g and i filters respectively.

@@

An arc-lamp image is declared unacceptable using the first 4 tests as employed
in tests of the flat-field images.  In addition, an arc-lamp exposure
is declared unacceptable ("BAD") if any of these following criteria are met:
5) not all 8 arc-lamps are on, 4 Neon lamps and 4 Mercury-Cadmium lamps,
6) the initial cross-correlation coefficient is less than 0.5,
7) fewer than 6 arc lines are acceptable after quality assurance checks.

The width of unblended arc lines are determined as an average over
each bundle of 20 spectra.  In short, the arc-lamp emission lines 
are extracted (with a Gaussian spectral profile) 
allowing free parameters for the area and width of each suitable line.
The resulting Gaussian widths per fiber bundle are fit 
with a low order polynomial as a function of average line position
(the fit is quadratic and cubic for the blue and red side respectively).
The fitted quantity is referred to as the arc-lamp dispersion,
but the values reflect the standard deviation of a Gaussian 
profile in pixel units and represents the line-spread function of the 
arc-lamp spectra.  Detailed analysis of the measured arc-line widths
compared to the observed sky emission line shapes show less than a 1\%
difference on average (McDonald et al. 2004).

@@

The accuracy of the SDSS spectral sky subtraction can been tested by examining
the residuals in final reduced spectra, where spectra of blank sky, high-z QSOs
and luminous red galaxies have been investigated to date.
For early releases of spectroscopic data, reduced by version 4 and include
data releases 4 and earlier, there have been mulitple confirmations of an
underestimation of the true variance as reported in the output of the 
spectroscopic pipeline.  By comparing multiple observations of the same
high-z QSOs in studies of the Lyman-$\alpha$ forest, the average variance was
shown to be underestimated by approximately 16\% (McDonald et al. 2004).  
A fraction of the effect arises due to large tails in the distribution 
of the noise and the remainder from the predicted FWHM of the noise 
distribution.  Switzer (2004) identified two causes for the underestimates,
citing both the awkward treatment of large sky residuals relative to the
supersky and the corresponding correction described above, and also a general
overestimation of the gains applied for the 8 CCD amplifiers.

Using the sample of plates here, we investigate the distribution of the 
noise in a large sample of sky fibers as
well as a sample of model fit luminous red galaxies.  Inclusion of
the latter sample is motivated by earlier studies which showed 
the spectroscopic residuals about the best fit galaxy models 
can be described by a sum of Gaussian and Laplace 
distributions (Bolton et al. 2004).  We construct histograms of the flux
pixels directly from the SDSS spectroscopic products for two versions
of the software (versions 4 \& 5).  In the first sample, flux measurements
from the sky fibers are shown relative to zero and normalized by their reported
inverse variances 
(${\rm residual} = {\rm flux} \, \sqrt{\rm inverse variance}$).  
In the second sample, residuals in galaxy spectra, 
which were targetted as luminous red galaxies, were constructed by subtracting
the best fit galaxy model and then normalizing by the inverse variance.

@@

We select relatively bright F dwarfs as spectrophotometric 
standards.  Ideally, we want the low-metallicity, halo stars in order 
to have spectra that have atmospheric opacities that are easy to model.
The F stars are chosen to satisfy the following color criteria:
$$ 0.6 < u-g < 1.2 $$
$$ 0.0 < g-r < 0.6 $$
$$ g-r > 0.75 * (u-g) - 0.45 $$
as shown in Figure ???.
This selects stars that are on the blue edge of the main
sequence in (u-g) color.  At high Galactic latitudes,
??? stars per square degree fall in this color box.
Targetting priority is given to
those stars with colors closest to the following:
$$ u-g = 0.80 $$
$$ g-r = 0.30 $$
$$ r-i = 0.10 $$
This is meant to select those stars closest in color to BD +17 4708,
which is computed to have a similar color (0.93, 0.28, 0.01, courtesy
of David Hogg, priv. comm.).
From these stars, we select 8 on each plate in the magnitude
range $15.5 < g < 17.0$ (which we call SPECTROPHOTO\_STD stars)
and 8 in the magnitude range $17.0 < g < 18.5$ (which we call
REDDEN\_STD stars).
These magnitude limits are chosen to avoid nearing saturation
of the CCDs on the bright end, but to still ensure good S/N
on the faint end.
Assuming an absolute magnitude of 4.3 for these stars, this
implies distances of $1.7 < d < 3.5$ kpc for the SPECTROPHOTO\_STD
stars and $3.5 < d < 6.9$ kpc for the REDDEN\_STD stars.
Note that the color-selection and magnitude cuts are done on
``instrinsic'' magnitudes, meaning that we first apply corrections
for SFD extinction under the assumption that these stars are
completely behind the Galactic dust.

@@

@@

\begin{figure}
\figurenum{11}
\epsscale{1.0}
\begin{figure}
\figurenum{11}
\epsscale{1.0}
\plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}
\figcaption{ Distribution of SDSS flux residuals normalized 
by the reported pipeline inverse variances.  The left plot shows
the distribution as a histrogram of 214 million sky pixels in 
reductions from both version 4 (DR4 and earlier, upper) 
and version 5 (lower) of spectro\_2d.  
The best fit functional form is shown as a solid red line (see text).
The filled region shows the area between the observed version 5 distribution 
of sky flux residuals and a gaussian with identical 1-sigma confidence limits.
The right plot is the same but shows the distribution of approximately 
550 million pixels drawn from residuals with respect to best fit models of
luminous red galaxies.
\label{fig11}
}
\end{figure}

>>>>

%------------------------------------------------------------------------------
% SUBJECT HEADINGS
%------------------------------------------------------------------------------
\emph{Subject headings: }
instrumentation: spectrographs --- techniques: spectroscopic --- 
methods: data analysis --- surveys -- galaxies: distances and redshifts
\end{abstract}

@@

In this paper, we describe the details of the SDSS 2-d spectroscopic pipeline
(idlspec2d) including steps starting with processing of the raw CCD frames 
to final flux calibration and spectral co-addition.  Downstream pipelines 
which calculate redshifts, spectral types, emission and absorption line 
identification, and velocity dispersions are described elsewhere.
As of writing the 2-d pipeline has been run successfully on all spectroscopic
data obtained in the course of the survey from February 28, 2000 
(Modified Julian Data, MJD 2451602) 
to June 15, 2005 (MJD 2453536).  Over this timespan, which includes public data 
releases EDR through DR5 (Stoughton et al 2002, 
Abazajian et al 2003, 2004, 2005, Adelman-McCarthy et al 2006)
, SDSS obtained survey quality spectroscopic data on
1656 fiber pluggings (640 fibers each in 1575 unique plates).   
A total of 972,111 sky positions were targetted, with the following 
breakdown in the best matched spectral classification:  651920 galaxies, 
208054 stars, 112137 QSOs and 51300 sky spectra.

Mean exposure time per plate is 3914 seconds, and the median is 3302 seconds.
;
;   Should we include a histogram of exposure times??
;

In Figure 1 we show three of the data quality diagnostics. This figure shows
the effective signal-to-noise ratio of each spectroscopic plug plate 
normalized to 3600 seconds exposure time, the median seeing in arcseconds,
and the RMS guide probe offset
in arcseconds, versus time in days after Feb 26, 2000 (MJD 2451600).

@@

In Figure 1, we show a top-view schematic of one the spectrographs.  Here
we provide a short summary of the light path starting at the slithead,
further details can be found in Uomoto et al. (20??) and Gunn et al. (2006).
The output light from the fiber is collimated by a spherical mirror
(of radius of curvature 1264 mm), which is aligned by 3 motor mounts to
control focus, tip and tilt.  A left/right Hartmann mask 
covers the collimator to act as both a spectrograph shutter and traditional 
Shack-Hartmann mask for focus tests.  The collimator in spectrograph 2
was coated with an enhanced silver coating and delivers slightly higher
reflectivity compared to the aluminum coated collimator in spectrograph 1.
The collimated beam encounters a 40mm thick dichroic (beamsplitter) which
reflects light blueward of 6000\AA.  The blue reflected light is dispersed
with a 640 l/mm grism with zero deviation at 4960\AA.  The red transmitted
light is dispersed with a 440 l/mm grism with zero deviation at 7400\AA.
The dispersed light in each arm of the spectrograph enters a transmission camera
with a 240mm focal length (demagnification of 2.5) which is tuned for the 
corresponding range of wavelengths.  Each of the 4 cameras produce RMS 
spot sizes of approximately 20$\mu m$ over a 16.5$\degree$ diameter 
field-of-view delivered to a single thinned Tektronix 2048x2048x24$\mu m$ CCD
(see Gunn et al. 2006).
An external focus ring is adjustable by hand for each CCD camera to accomodate
differential focus changes not corrected with the common collimator.
The four spectrograph cameras are labeled $``$b1, r1, b2, r2", where
1 and 2 refer to the spectrographs and b and r to their blue and red sides.

@@

A plate is mounted in the telescope through a efficient procedure described
by Gunn et al (2006).  The switch from the imaging camera to a spectroscopic
cartridge takes an average of 30 minutes, while a switch from one cartridge
to another takes 10 minutes.  Once the cartridge is mounted, 
the telescope is slewed to the coordinates of the plate center.  
Before acquisition of the guide stars, the 8 flat-field petals are closed and
a sequence of one 10-second flat-field exposure and one 2-second 
Neon-Mercury-Cadmium arc exposure are obtained.  While the arc-lamp frame is
being read out, the flat-field petals are opened, and guide stars are acquired
through the 11 coherent guide bundles (9 are 7$"$ in diameter and 
2 are even larger at 11$"$).
With at least 4 guide stars acquired, closed loop guiding is started, and
usually converges to stable guiding within 2 minutes.
The plate scale of the SDSS telescope can be adjusted, and the guiding software
suggests a scale change if one is required.  Once close loop 
guiding has settled, the first science exposure is started.  The typical 
exposure time is 900 seconds, but exposure times from 600 to 1200 second are
not unexpected based on conditions and time constraints.  A second exposure
is started immediately after the first science exposure is complete and read
out.
A fast Linux workstation onsite copies the data from the data acquisition host,
and automatically performs a quick spectral extraction with a subset of the 
steps explained here.  The results are posted to sos.apo.nmsu.edu, and
the SDSS observers can check the signal-to-noise ratio (S/N) 
obtained in each of the 4 spectroscopic
cameras in real-time.  A plate is considered to have met SDSS survey standards
if all 4 cameras have a cumulative (S/N)$^2 > 15$ at 
fiber magnitudes of 20.33 and 20.06 in SDSS g and i filters respectively.

\section{IMAGE PROCESSING}
\label{sec_process}

sdssproc:

1000 lines of IDL code...

read opEcalib
     op something else

Read file header, require...
mention spHdrFix here.

\section{SPATIAL TRACING}
\label{sec_tracing}

trace320, Trace crude, and trace fitting

@@

An arc-lamp image is declared unacceptable using the first 4 tests as employed
in tests of the flat-field images.  In addition, an arc-lamp exposure
is declared unacceptable (``BAD") if any of these following criteria are met:
5) not all 8 arc-lamps are on: 4 Neon lamps and 4 Mercury-Cadmium lamps,
6) the initial cross-correlation coefficient is less than 0.5,
7) fewer than 6 arc lines are acceptable after quality assurance checks.

The width of unblended arc lines are determined as an average over
each bundle of 20 spectra.  In short, the arc-lamp emission lines 
are extracted (with a Gaussian spectral profile) 
allowing free parameters for the area and width of each suitable line.
The resulting Gaussian widths per fiber bundle are fit 
with a low order polynomial as a function of average line position
(the fit is quadratic and cubic for the blue and red side respectively).
The fitted quantity is referred to as the arc-lamp dispersion,
but the values reflect the standard deviation of a Gaussian 
profile in pixel units and represents the line-spread function of the 
arc-lamp spectra.  Detailed analysis of the measured arc-line widths
compared to the observed sky emission line shapes show less than a 1\%
difference on average (McDonald et al. 2006).

@@

The accuracy of the SDSS spectral sky subtraction can been tested by examining
the residuals in final reduced spectra, where spectra of blank sky, high-z QSOs
and luminous red galaxies have been investigated to date.
For early releases of spectroscopic data, reduced by version 4 and include
data releases 4 and earlier, there have been mulitple confirmations of an
underestimation of the true variance as reported in the output of the 
spectroscopic pipeline.  By comparing multiple observations of the same
high-z QSOs in studies of the Lyman-$\alpha$ forest, the average variance was
shown to be underestimated by approximately 16\% (McDonald et al. 2006).  
A fraction of the effect arises due to large tails in the distribution 
of the noise and the remainder from the predicted FWHM of the noise 
distribution.  Switzer (2004) identified two causes for the underestimates,
citing both the awkward treatment of large sky residuals relative to the
supersky and the corresponding correction described above, and also a general
overestimation of the gains applied for the 8 CCD amplifiers.

Using the sample of plates here, we investigate the distribution of the 
noise in a large sample of sky fibers as
well as a sample of model fit luminous red galaxies.  Inclusion of
the latter sample is motivated by earlier studies which showed 
the spectroscopic residuals about the best fit galaxy models 
can be described by a sum of Gaussian and Laplace 
distributions (Bolton et al. 2004).  We construct histograms of the flux
pixels directly from the SDSS spectroscopic products for two versions
of the software (versions 4 \& 5).  In the first sample, flux measurements
from the sky fibers are shown relative to zero and normalized by their reported
inverse variances 
(${\rm residual} = {\rm flux} \, \sqrt{\rm 1}$).  
In the second sample, residuals in galaxy spectra, 
which were targetted as luminous red galaxies, were constructed by subtracting
the best fit galaxy model and then normalizing by the inverse variance.

@@

We select relatively bright F dwarfs as spectrophotometric 
standards.  Ideally, we want the low-metallicity, halo stars in order 
to have spectra that have atmospheric opacities that are easy to model.
The F stars are chosen to satisfy the following color criteria:
$$ 0.6 < u-g < 1.2 $$
$$ 0.0 < g-r < 0.6 $$
$$ g-r > 0.75 * (u-g) - 0.45 $$
as shown in Figure ???.
This selects stars that are on the blue edge of the main
sequence in (u-g) color.  At high Galactic latitudes,
??? stars per square degree fall in this color box.
Targetting priority is given to
those stars with colors closest to the following:
$$ u-g = 0.80 $$
$$ g-r = 0.30 $$
$$ r-i = 0.10 $$
in order to select those stars closest in color to BD +17 4708 
(0.93, 0.28, 0.01, courtesy of David Hogg, priv. comm.).
From these stars, we select 8 on each plate in the magnitude
range $15.5 < g < 17.0$ (which we call SPECTROPHOTO\_STD stars)
and 8 in the magnitude range $17.0 < g < 18.5$ (which we call
REDDEN\_STD stars).
These magnitude limits are chosen to avoid nearing saturation
of the CCDs on the bright end, but to still ensure good S/N
on the faint end.
Assuming an absolute magnitude of 4.3 for these stars, this
implies distances of $1.7 < d < 3.5$ kpc for the SPECTROPHOTO\_STD
stars and $3.5 < d < 6.9$ kpc for the REDDEN\_STD stars.
Note that the color-selection and magnitude cuts are done on
``instrinsic'' magnitudes, meaning that we first apply corrections
for SFD extinction under the assumption that these stars are
completely behind the Galactic dust.

@@

\bibitem[McDonald et al.(2006)]{2006ApJS..163...80M} McDonald, P., et al.\ 
2006, \apjs, 163, 80

\bibitem[Stoughton et al.(2002)]{2002AJ....123..485S} Stoughton, C., et 
al.\ 2002, \aj, 123, 485

\bibitem[Adelman-McCarthy et al.(2006)]{2006ApJS..162...38A} 
Adelman-McCarthy, J.~K., et al.\ 2006, \apjs, 162, 38

\bibitem[Abazajian et al.(2005)]{2005AJ....129.1755A} Abazajian, K., et 
al.\ 2005, \aj, 129, 1755

\bibitem[Abazajian et al.(2004)]{2004AJ....128..502A} Abazajian, K., et 
al.\ 2004, \aj, 128, 502

\bibitem[Abazajian et al.(2003)]{2003AJ....126.2081A} Abazajian, K., et 
al.\ 2003, \aj, 126, 2081

@@

\begin{figure}
\figurenum{11}
\epsscale{1.0}
\plottwo{plot_sky_reshist.ps}{plot_lrg_reshist.ps}
\figcaption{ Distribution of SDSS flux residuals normalized 
by the reported pipeline inverse variances.  The left plot shows
the distribution as a histrogram of 214 million sky pixels in 
reductions from both version 4 (DR4 and earlier, upper) 
and version 5 (lower) of spectro\_2d.  
The best fit functional form is shown as a solid red line (see text).
The filled region shows the area between the observed version 5 distribution 
of sky flux residuals and a gaussian with identical 1-sigma confidence limits.
The right plot is the same but shows the distribution of approximately 
550 million pixels drawn from residuals with respect to best fit models of
luminous red galaxies.
\label{fig11}
}
\end{figure}
